---
title: "Script"
author: "Sarah Roberts"
date: "9/14/2020"
output: html_document
---

The purpose of this script is to help Crystal and Bea with their MP projects. In this script I will start from the dataset that I created called total.csv. To create total, follow the steps here: https://github.com/sr197/Sticky_Fish. That script requires the user to download NEFSC bottom trawl survey data from Ocean Adapt. https://oceanadapt.rutgers.edu/ . 

This script does the following: 
1. Explore the dataset including species collected, environmental variables, spatial distribution. 
2. Run species distribution models for presence absence and biomass. 

note I run this for the total dataset, but these species behave very differently in the spring and fall so it would be good to split that up. 
#packages 
```{r}
library(tidyverse) #for data manipulation 
library(dplyr) #for data manipulation 
library(mgcv) #for running gams 
#spatial 
library(sp) #mean centroid
library(spatialEco)#mean centroid
library(geosphere)#mean centroid
library(rgdal)#kernel density 
library(adehabitatHR)#kernel density 
library(ks)#kernel density 
library(sf)#kernel density 
library(raster) #kernel density projection 
#plotting 
library(ggplot2) 
library(ggpmisc)
library(reshape)
library(ggpubr)
library(ggsci)
library(scales)
library(rnaturalearth) #gives you some nice shapefiles 

```

#1 Explore data 
```{r}
total <- read.csv("total.csv")
View(total)
colnames(total)
```
I am going to make a quick version of sediment that is not a factor so instead it is sediment size. 
```{r}
total$SEDSIZE <- ifelse(total$SEDIMENT ==  '0 - 0.03 Silt/Mud', .03, 
                        ifelse(total$SEDIMENT ==  '0.03 - 0.17 Sand', .17,
                               ifelse(total$SEDIMENT ==  '0.17 - 0.35 Sand',.35, 
                                      ifelse(total$SEDIMENT ==  '0.35 - 0.36 Sand', .36,
                                             ifelse(total$SEDIMENT ==  '0.36 - 0.48 Sand', .48, ifelse(total$SEDIMENT ==  '0.48+ Coarse Sand to Gravel',.50,NA))))))
```
##1.1 Atlantic cod 
Lets just select out cod and some other env variables
```{r}
cod <- total %>% dplyr::select(GADUS.MORHUA, year, date, lat, lon, depth, SEDIMENT, SEDSIZE, bottemp, surfsalin, season)
```

lets explore how many cod there are 
```{r}
sum(cod$GADUS.MORHUA>0)
#they are present in 4000 tows, that seems like a lot
```

lets make a dataset of presence absence 
```{r}
cod_pres <- cod
cod_pres$GADUS.MORHUA[cod_pres$GADUS.MORHUA > 0] <- 1 


cod_pres$GADUS.MORHUA[cod_pres$GADUS.MORHUA=="0"] <- "absent"
cod_pres$GADUS.MORHUA[cod_pres$GADUS.MORHUA=="1"] <- "present"
```

lets plot the data 
```{r}

world <- ne_countries(scale = "medium", returnclass = "sf")

ggplot(data = world) +
     geom_sf() +
    geom_point(data = cod_pres, aes(x = lon, y = lat, col = GADUS.MORHUA)) +
    coord_sf(xlim=c(-90, -60), ylim=c(30,50), expand = FALSE)

ggplot(data = world) +
     geom_sf() +
    geom_point(data = subset(cod, cod$GADUS.MORHUA>0), aes(x = lon, y = lat, col = GADUS.MORHUA)) + scale_colour_gradientn(colors = c("#fff7f3", "#fde0dd","#fcc5c0", "#fa9fb5", "#f768a1","#dd3497","#ae017e","#7a0177","#49006a"), values = rescale(c( .01, .05, .1, 1, 50))) +
    coord_sf(xlim=c(-90, -60), ylim=c(30,50), expand = FALSE)
```


#2 Modeling 
I can think of two questions related to cod. 
1. How does climate and (insert other variable like fishing pressure) influence the biomass of cod. Which variables influence cods biomass more? 
2. How does climate and x influence the spatial distribution of cod? 
  This question is going to need to look into calculating measurements of distribution (i.e range size, mean centroid, etc). This question is based on analysis like Nye et al 
  https://www.researchgate.net/publication/230881627_Changing_spatial_distribution_of_fish_stocks_in_relation_to_climate_and_population_size_on_the_Northeast_United_Sates_continental_shelf 
  
##2. 1 SDM Question 1
You should be familiar with this from stats classes and gis. Basically you want to follow these steps. 
1. create a species distribution model 
1.1. split into training and testing datasets, use cross validation to explore model accuracy etc. This may involve the caret package or AIC to land on the "best" model. 
2. predict model on new data if you want to. 



```{r}
#lets run a simple glm on the presence absence and cpue dataset
# logistic regression:
cod_pres <- cod
cod_pres$GADUS.MORHUA[cod_pres$GADUS.MORHUA > 0] <- 1 


gm1 <- glm(GADUS.MORHUA ~ bottemp + depth + SEDSIZE + surfsalin, 
            family = binomial(link = "logit"), data=cod_pres)

summary(gm1)

gm2 <- glm(GADUS.MORHUA ~ bottemp + depth + SEDSIZE + surfsalin, data=cod)

summary(gm2)
```
notice the difference in significance. 

I like random forests so lets also try that 
```{r}
library(randomForest)
cod_pres$GADUS.MORHUA <-factor(ifelse(cod$GADUS.MORHUA==1, "present", "absent"), levels = c("present", "absent"))
model <- factor(GADUS.MORHUA) ~ bottemp + depth + SEDSIZE + surfsalin

rf1 <- randomForest(model, data=cod_pres,na.action = na.omit)
importance(rf1)
varImp(rf1)
varImpPlot(rf1)


model <- GADUS.MORHUA ~ bottemp + depth + SEDSIZE + surfsalin

rf2 <- randomForest(model, data=cod,na.action = na.omit)
importance(rf2)
varImp(rf2)
varImpPlot(rf2)
```
kind of strange that depth is important for the presence of cod, but not for the biomass (or abundance of cod). Thinking about that critically is going to be important for how we model them. 

probably should have plotted this earlier but it might be helpful 
```{r}
plot(cod$depth, cod$GADUS.MORHUA)


ggplot(data=cod_pres, aes(x=GADUS.MORHUA, y=depth)) +
  geom_boxplot()+ 
  stat_summary(fun.y=mean, colour="darkred", geom="point", 
               shape=18, size=3,show_guide = FALSE)  +
    labs(y = "depth", x = "presence")+ 
  theme_classic()


ggplot(data=cod_pres, aes(x=GADUS.MORHUA, y=bottemp)) +
  geom_boxplot()+ 
  stat_summary(fun.y=mean, colour="darkred", geom="point", 
               shape=18, size=3,show_guide = FALSE)  +
    labs(y = "bt", x = "presence")+ 
  theme_classic()

ggplot(data=cod_pres, aes(x=GADUS.MORHUA, y=surfsalin)) +
  geom_boxplot()+ 
  stat_summary(fun.y=mean, colour="darkred", geom="point", 
               shape=18, size=3,show_guide = FALSE)  +
    labs(y = "salinity", x = "presence")+ 
  theme_classic()

ggplot(data=cod_pres, aes(x=GADUS.MORHUA, y=SEDSIZE)) +
  geom_boxplot()+ 
  stat_summary(fun.y=mean, colour="darkred", geom="point", 
               shape=18, size=3,show_guide = FALSE)  +
    labs(y = "sediment size", x = "presence")+ 
  theme_classic()



```


###2.1.2 step AIC 
refer to marine gis to do this

###2.1.3 cross validation 
I encourage you to explore the caret package here that allows for simple cross validation. 
caret package tutorial: http://topepo.github.io/caret/index.html 

```{r}
cod_pres$GADUS.MORHUA <-factor(ifelse(cod$GADUS.MORHUA==1, "present", "absent"), levels = c("present", "absent"))
library(caret)

fitControl = trainControl(method = "cv", repeats = 5,
                                       classProbs = TRUE,
                                       summaryFunction = twoClassSummary, verbose = FALSE)

model <- GADUS.MORHUA ~ bottemp + depth + SEDSIZE + surfsalin


glm3 <- train(model, data=cod_pres, method="glm", metric = "ROC",
              maximize = FALSE, trControl = fitControl, na.action = na.omit)
glm3
summary(glm3)

```


##2.2 SDM Question 2. 

How does climate and x influence the spatial distribution of cod? 

###2.2.1 Spatial Metrics 
Mean centroid, mean latitude, mean depth etc. 
```{r}
Mode <- function(x) {
 ux <- unique(x)
 ux[which.max(tabulate(match(x, ux)))]
}
# weighted mean for use with summarize(). values in col 1, weights in col 2
wgtmean = function(x, na.rm=FALSE) {questionr::wtd.mean(x=x[,1], weights=x[,2], na.rm=na.rm)}
wgtse = function(x, na.rm=TRUE){ 
  if(sum(!is.na(x[,1]) & !is.na(x[,2]))>1){
    if(na.rm){
      return(sqrt(wtd.var(x=x[,1], weights=x[,2], na.rm=TRUE, normwt=TRUE))/sqrt(sum(!is.na(x[,1] & !is.na(x[,2])))))
    } else {
      return(sqrt(wtd.var(x=x[,1], weights=x[,2], na.rm=FALSE, normwt=TRUE))/sqrt(length(x))) # may choke on wtd.var without removing NAs
    }
  } else {
    return(NA) # NA if vector doesn't have at least 2 values
  }
}
se <- function(x) sd(x)/sqrt(length(x)) # assumes no NAs
sumna <- function(x){
  #acts like sum(na.rm=T) but returns NA if all are NA
  if(!all(is.na(x))) return(sum(x, na.rm=T))
  if(all(is.na(x))) return(NA)
}
meanna = function(x){
  if(!all(is.na(x))) return(mean(x, na.rm=T))
  if(all(is.na(x))) return(NA)
}

```


```{r}
# Calculate mean position through time for species 
## Calculate mean latitude and depth of each species by year within each survey/region
### mean lat/lon/depth for each stratum

ocean_adapt <- read.csv("neus_done_by_sarah.csv")


dat_strat <- ocean_adapt %>% 
  dplyr::select(stratum, lat, lon, depth, stratumarea, haulid) %>% 
  distinct(stratum, haulid, .keep_all = T) %>% 
  group_by(stratum) %>% 
  summarise(lat = meanna(lat), 
            lon = meanna(lon), 
            depth = meanna(depth), 
            stratumarea = meanna(stratumarea))

### mean wtcpue in each stratum/yr/spp (new code includes more lines because it
### includes rows that do not have a common name)
dat_strat_yr <- ocean_adapt %>% 
  group_by(spp, stratum, year) %>% 
  summarise(wtcpue = meanna(wtcpue))

# add stratum lat/lon/depth/area
dat_strat_yr <- left_join(dat_strat_yr, dat_strat, by = c("stratum"))

# index of biomass per stratum: mean wtcpue times area
dat_strat_yr <- dat_strat_yr %>% 
  mutate(wttot = wtcpue * stratumarea)

# calculate mean lat
cent_bio_lat <- dat_strat_yr %>% 
  group_by(spp, year) %>% 
  summarise(lat = questionr::wtd.mean(lat, wttot, na.rm = T))

# mean depth
cent_bio_depth <- dat_strat_yr %>% 
  group_by(spp, year) %>% 
  summarise(depth = questionr::wtd.mean(depth, wttot, na.rm = T))

# mean lon
cent_bio_lon <- dat_strat_yr %>% 
  group_by(spp, year) %>% 
  summarise(lon = questionr::wtd.mean(lon, wttot, na.rm = T))


# merge
cent_bio <- left_join(cent_bio_lat, cent_bio_depth, by = c( "spp","year"))
cent_bio <- left_join(cent_bio, cent_bio_lon, by = c("spp", "year"))

# standard error for lat
cent_bio_lat_se <- dat_strat_yr %>%
  group_by(spp, year) %>% 
  summarise(lat_se = sqrt(questionr::wtd.var(lat, wttot, na.rm=TRUE, normwt=TRUE))/sqrt(sum(!is.na(lat) & !is.na(wttot))))

cent_bio <- left_join(cent_bio, cent_bio_lat_se, by = c("spp", "year"))

cent_bio_depth_se <- dat_strat_yr %>%
  group_by(spp, year) %>% 
  summarise(depth_se = sqrt(questionr::wtd.var(depth, wttot, na.rm=TRUE, normwt=TRUE))/sqrt(sum(!is.na(depth) & !is.na(wttot))))

cent_bio <- left_join(cent_bio, cent_bio_depth_se, by = c("spp", "year"))

cent_bio_lon_se <- dat_strat_yr %>%
  group_by(spp, year) %>% 
  summarise(lon_se = sqrt(questionr::wtd.var(lon, wttot, na.rm=TRUE, normwt=TRUE))/sqrt(sum(!is.na(lon) & !is.na(wttot))))

cent_bio <- left_join(cent_bio, cent_bio_lon_se, by = c("spp", "year"))

BY_SPECIES_DATA <- cent_bio %>%
  ungroup() %>% 
  arrange(spp, year)

write.csv(cent_bio, "cent_bio.csv")


```


Then you could go and run a glm but instead of biomass as the response variable it would be annual cent_bio latitude or something. This would require annual predictor variables as well. 

```{r}
cent_cod <- cent_bio %>% filter(spp == "GADUS MORHUA")

ggplot(data = world) +
     geom_sf() +
    geom_point(data = cent_cod, aes(x = lon, y = lat, col = year)) + scale_colour_gradientn(colors = c("#fff7f3", "#fde0dd","#fcc5c0", "#fa9fb5", "#f768a1","#dd3497","#ae017e","#7a0177","#49006a")) +
    coord_sf(xlim=c(-90, -60), ylim=c(30,50), expand = FALSE)

cent_croak <- cent_bio %>% filter(spp == "MICROPOGONIAS UNDULATUS")

ggplot(data = world) +
     geom_sf() +
    geom_point(data = cent_croak, aes(x = lon, y = lat, col = year)) + scale_colour_gradientn(colors = c("#fff7f3", "#fde0dd","#fcc5c0", "#fa9fb5", "#f768a1","#dd3497","#ae017e","#7a0177","#49006a")) +
    coord_sf(xlim=c(-90, -60), ylim=c(30,50), expand = FALSE)



```

```{r}
library(reshape2)
library(ggpubr)
dat.m <- melt(cent_cod,id.vars='year', measure.vars=c(
"lat", "depth", "lon"))

ggplot(dat.m, aes(x=year, y=value)) +
    geom_point(shape=1) +    # Use hollow circles
    geom_smooth(method=lm) +
  facet_wrap(~variable, scales = "free")

dat.m <- melt(cent_croak,id.vars='year', measure.vars=c(
"lat", "depth", "lon"))

ggplot(dat.m, aes(x=year, y=value)) +
    geom_point(shape=1) +    # Use hollow circles
    geom_smooth(method=lm) +
  facet_wrap(~variable, scales = "free")


```

so croaker have moved offshore and north while cod have gone offshore and deeper. cool!